---
title: "Sea Ice test"
output:
  html_document:
    df_print: paged
---

```{r libraries, echo=T, results='hide', message=F, warning=F}
library(rstan)
library(gdata)
library(bayesplot)
```


Read in file and drop one row with no data:

```{r readData}
seaIce <- read.csv("iceData.csv")
seaIce <- seaIce[-c(10), ]
```

Look at data
```{r look}
head(seaIce)
plot(extent ~ year, data = seaIce)
```

Now run a general linear model and add line to plot
```{r linModel}
lm1 <- lm(extent ~ year, data = seaIce)
summary(lm1)
plot(extent ~ year, data = seaIce)
abline(lm1)

```

Reformat data to start at year 0 and rerun linear model
```{r reformat}

extent <- seaIce$extent
year <- I(seaIce$year - 1978)
N <- length(seaIce$year)

stan_data <- list(N = N, year = year, extent = extent)

lm1 <- lm(extent ~ year)
summary(lm1)

lm_alpha <- summary(lm1)$coeff[1]  # the intercept
lm_beta <- summary(lm1)$coeff[2]  # the slope
lm_sigma <- sigma(lm1)  # the residual error

```

Load in an run stan model
```{r runStan}
stan_model <- "seaIceModel.stan"
fit <- stan(file = stan_model, data = stan_data, warmup = 500, iter = 1000, chains = 4, cores = 2, thin = 1)
```

Examine fit
```{r examFit}
fit

posterior <- extract(fit)
str(posterior)
```

Plot comparing to earlier linear prediction. Black dashes is linear, red line is stan.

```{r compLinStan}
plot(extent ~ year, pch = 20)

abline(lm1, col = 1, lty = 2, lw = 3)
abline(mean(posterior$alpha), mean(posterior$beta), col = 2, lw = 1)
```
These are identical because it is a very simple linear model with only weakly informative parameters.

You can visualize the variability by plotting multiple estimates from the posterior.

```{r variability}
plot(extent ~ year, pch = 20)

for (i in 1:500) {
 abline(posterior$alpha[i], posterior$beta[i], col = "gray", lty = 1)
}

abline(mean(posterior$alpha), mean(posterior$beta), col = 2, lw = 2)
```

Look at traceplots to see convergence. Can do this either directly referencing the posterior or by calling traceplot.

```{r traceplots}
par(mfrow = c(1,3))
plot(posterior$alpha, type = "l")
plot(posterior$beta, type = "l")
plot(posterior$sigma, type = "l")

traceplot(fit,pars=c("alpha","beta","sigma"))
```

Plot the posterior distributions, as well as the values from the linear fit to compare. Can do either by referencing posterior or by using stan_dens().

```{r posterior}
par(mfrow = c(1,3))

plot(density(posterior$alpha), main = "Alpha")
abline(v = lm_alpha, col = 4, lty = 2)

plot(density(posterior$beta), main = "Beta")
abline(v = lm_beta, col = 4, lty = 2)

plot(density(posterior$sigma), main = "Sigma")
abline(v = lm_sigma, col = 4, lty = 2)

stan_dens(fit,pars=c("alpha","beta","sigma"))
```

Can also calculate the probability that a certain parameter is above or below some value. In this case the probability that ice has been increasing.

```{r chance}
sum(posterior$beta>0)/length(posterior$beta)
```

Can plot values and credible intervals for parameters. In this case the 95% intervals are so small that you can't see them.

```{r paramValues}
plot(fit,pars=c("alpha","beta","sigma"),show_density = FALSE, ci_level = 0.5, outer_level = 0.95, fill_color = "salmon")
```

In the generated quantities section I had stan generate predicted values using rng for each data point.

```{r predictVals}
ext_rep <- as.matrix(fit, pars = "ext_rep")
dim(ext_rep)
```

Can compare density of y with density of predicted values
```{r density}
ppc_dens_overlay(extent, ext_rep[1:200, ])
```

